---
title: "Exploring Springleaf Data"
output: html_document
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(caret)
library(readr)
library(xgboost)
library(ggplot2)
library(R.utils)
library(gridExtra)
library(lubridate)
library(data.table)
library(Matrix)
require(plyr)
require(Hmisc)
library(maps)
library(maptools)
library(sp)
library(corrplot)
```

Read in the Data and remove ID and response variables

```{r warning=FALSE, message=FALSE}

train <- read_csv("D:/kaggle/Springleaf/DATA/CSV/train.csv")

test <- read_csv("D:/kaggle/Springleaf/DATA/CSV/test.csv")

dim(train); dim(test)

response <- train$target

train_ID <- train$ID

test_ID <- test$ID

```

Modify train and test sets

```{r warning=FALSE, message=FALSE}

training <- subset(train, select = -c(ID, target))

testing <- subset(test, select = -c(ID))

dim(training); dim(testing)
```

CHECKING FOR DUPLICATE ROWS

```{r warning=FALSE, message=FALSE}

nrow(training) - nrow(unique(training)) #DONT RUN TAKES RIDICULOUSLY LONG TIME

```
CHECKING FOR UNIQUE ELEMENTS IN A COLUMN

```{r warning=FALSE, message=FALSE}

col_unique <- sapply(training, function(x) unique(x))

```

CHECKING FOR NUMBER OF UNIQUE VALUES IN A COLUMN

```{r warning=FALSE, message=FALSE}

col_ele <- as.data.frame(lapply(training, function(x) length(unique(x))))

##GET THE COLUMN NAMES

length(col_ele)

```

CHECK FOR COLUMNS WITH 1,2,3 UNIQUE ELEMENTS

```{r warning=FALSE, message=FALSE}

length(col_ele[col_ele == seq(1,3,1)])

length(col_ele[col_ele == 1])

length(col_ele[col_ele == 2]) #MOSTLY COLUMNS WITH DUMMY VARS

length(col_ele[col_ele == 3])

```

CHECK COLUMNS WITH ONLY 1 UNIQUE VALUE

```{r warning=FALSE, message=FALSE}

unique_one <- subset(col_ele , select = c(col_ele == 1))
 
unique_oneDF <- training[, c(names(unique_one))]

head(unique_oneDF)
```


IDENTIFY AND SEPERATE NUMERIC AND NON NUMERIC COLUMNS
```{r warning=FALSE, message=FALSE}

training_num <- training[, sapply(training, is.numeric)] #CHECK WHETHER LAPPLY WORKS

training_char <- training[,sapply(training, is.character)]

cat("Number of Numerical columns :" , dim(training_num)[2], "|Number of character columns :", dim(training_char)[2])

```

LOOK INTO NUMERIC COLUMNS DF

CHECK FOR UNIQUE AND LENGTH OF UNIQUE COLUMNS

```{r warning=FALSE, message=FALSE}

str(lapply(training_num, unique)) 

str(lapply(training_num, function(x) length(unique(x)))) 
```

```{r warning=FALSE, message=FALSE}

numeric_ele <- (lapply(training_num, function(x) length(unique(x))))

#CHECK FOR COLUMNS WITH 1,2,3 UNIQUE ELEMENTS

length(numeric_ele[numeric_ele == 1])

length(numeric_ele[numeric_ele == 2]) #MOSTLY COLUMNS WITH DUMMY VARS

length(numeric_ele[numeric_ele == 3])

#CHECK COLUMNS WITH ONLY 1 UNIQUE VALUE

numeric_one <- subset(numeric_ele , subset  = c(numeric_ele == 1))

numeric_oneDF <- training_num[, c(names(numeric_one))]

#CHECK ELEMENTS

lapply(numeric_oneDF, table)

#CHECK COLUMNS WITH 2 UNIQUE VALUES

numeric_two <- subset(numeric_ele , subset = c(numeric_ele == 2))

numeric_twoDF <- training_num[, c(names(numeric_two))]

#CHECK ELEMENTS

lapply(numeric_twoDF, table)

```

LOOK INTO CHARACTER COLUMNS DF

```{r warning=FALSE, message=FALSE}

str(lapply(training_char, unique), vec.len =4 )

char_ele <- (lapply(training_char, function(x) length(unique(x))))

#CHECK FOR COLUMNS WITH 1,2,3 UNIQUE ELEMENTS

range(char_ele)

length(char_ele[char_ele == 1])

length(char_ele[char_ele == 2])

length(char_ele[char_ele == 3])

```

SOME COLUMNS HAVE NAMES: FIRST SEPERATE DATES THEN LOOK INTO THEM 

SEPERATE OUT DATES AND TIMES INTO  DIFFERENT DFS

```{r warning=FALSE, message=FALSE}

training_date <- training_char[, grep("JAN1|FEB1|MAR1", training_char)]

names(training_date)

```

```{r warning=FALSE, message=FALSE}

#REMOVE DATES FROM char DF

training_charD <- training_char[, !(names(training_char) %in% names(training_date))] 

#training_charD has only CHAR variables checking further

str(lapply(training_charD, unique), vec.len =4 )

```

SEPERATE FIELDS WITH BINARY VALUES

```{r warning=FALSE, message=FALSE}

charD_ele <- (lapply(training_charD, function(x) length(unique(x))))

charD_two <- subset(charD_ele, subset = c(charD_ele == 2))

names(charD_two)

charD_twoDF <- training_char[, c(names(charD_two))]

#CHECK ELEMENTS OF DF

lapply(charD_twoDF, table)

```

DRILL DOWN FURTHER BY REMOVING THESE COLS

```{r warning=FALSE, message=FALSE}

training_charD_edit <- training_charD[, !(names(training_charD) %in% names(charD_two))] 

#training_charD_edit has only CHAR variables checking further

str(lapply(training_charD_edit, unique), vec.len = 4)

#FURTHER DRILL DOWN BY SEPERATING NAMES

charD_many <- subset(charD_ele, subset = c(charD_ele == 1824 | 
                                             charD_ele == 609 | charD_ele == 12387))
names(charD_many)

##PLACE EACH OF THE NAMES IN A INDI.. DF AND CHECK IT OUT

charDF <- data.frame(cities = training_charD_edit$VAR_0200)

dim(charDF)

head(charDF)

nrow(na.omit(charDF))

charDF1 <- data.frame(small_work = training_charD_edit$VAR_0404)

length(charDF1[charDF1 != -1])

##REMOVING NA`S FROM CHARDF1

no_NA1 <- charDF1[charDF1 != -1]

no_NA1 <- no_NA1[no_NA1 != ""]

no_NA1 <- no_NA1[no_NA1 != "CONTACT"]

no_NA1 <- no_NA1[no_NA1 != "CONTA"]

NA1 <- (as.data.frame(table(no_NA1)))

charDF2 <- data.frame(big_work = training_charD_edit$VAR_0493)

length(charDF2[charDF2 != -1])

nrow(charDF2) - length(charDF2[charDF2 != -1])

no_NA2 <- charDF2[charDF2 != -1]

no_NA2 <- no_NA2[no_NA2 != ""]

NA2 <- (as.data.frame(table(no_NA2)))

NA1

NA2

```

DRILL DOWN INTO DATES VALUES

```{r warning=FALSE, message=FALSE}

names(training_date) 

training_date <- sapply(training_date, function(x) strptime(x, "%d%B%y :%H:%M:%S"))

training_date = do.call(cbind.data.frame, training_date)

head(training_date)

##DRILL DOWN TIME VALUES

training_time <- training_date[, names(training_date) %in% c("VAR_0204","VAR_0217")]

training_time <- data.frame(sapply(training_time, function(x) strftime(x, "%H:%M:%S")))

dim(training_time); head(training_time)

training_hour <- as.data.frame(sapply(training_time, function(x) 
as.numeric(as.character(substr(x, 1,2)))))

dim(training_hour); head(training_hour)

training_minute <- as.data.frame(sapply(training_time, function(x) 

  as.numeric(as.character(substr(x, 4,5)))))

dim(training_minute); head(training_minute)

```

HISTOGRAM OF DATES

```{r  warning=FALSE, message=FALSE,  fig.width = 8, fig.height = 6}

par(mar=c(2,2,2,2),mfrow=c(4,4))

for(i in 1:16) hist(training_date[,i], "weeks", format = "%d %b %y", main = 
                      colnames(training_date)[i], xlab="", ylab="")
```

HISTOGRAM OF TIMES

```{r  warning=FALSE, message=FALSE,  fig.width = 5, fig.height = 3}

par(mar=c(2,2,2,2),mfrow=c(1,2))

for(i in 1:2) hist(training_hour[,i], main = paste(colnames(training_hour)[i], "hourly"), 
                   breaks = c(0:24), xlab="", ylab="")
```

TAKING A LOOK AT THE GEOGRAPHICAL BREAKDOWN OF STATE FEATURES

```{r  warning=FALSE, message=FALSE,  fig.width = 10, fig.height = 5}

mapUSA <- map('state', fill=TRUE, plot=FALSE)

nms <- sapply(strsplit(mapUSA$names,  ':'),  function(x)x[1])

USApolygons <- map2SpatialPolygons(mapUSA,  IDs = nms,  CRS('+proj=longlat'))

mapStates = function(df, feat){

    dat = data.frame(table(df[,feat]))
  
    names(dat) = c("state.abb", "value")
  
    dat$states <- tolower(state.name[match(dat$state.abb,  state.abb)])
  
  idx <- match(unique(nms),  dat$states)
  
  dat2 <- data.frame(value = dat$value[idx], state = unique(nms))
  
  row.names(dat2) <- unique(nms) 
  
  USAsp <- SpatialPolygonsDataFrame(USApolygons,  data = dat2)
  
  spplot(USAsp['value'], main=paste(feat, "value count"), 
         col.regions=rev(heat.colors(21)))
}

grid.arrange(mapStates(training_char, "VAR_0274"), mapStates(training_char, "VAR_0237"),ncol=2)

```

NUMBER OF UNIQUE VALUES PER COLUMN

```{r  warning=FALSE, message=FALSE,  fig.width = 12, fig.height = 6}

num_unique = sapply(training_num, function(x) length(unique(x)))

char_unique = sapply(training_char, function(x) length(unique(x)))

date_unique = sapply(training_date, function(x) length(unique(x)))

all_unique = rbind(data.frame(count=num_unique, type="Numerical"), 
               
                   data.frame(count=char_unique, type="Character"), 
               
                   data.frame(count=date_unique, type="Date"))

```

PLOT OF NUMBER OF UNIQUE VALUES PER COLUMN

```{r  warning=FALSE, message=FALSE,  fig.width = 12, fig.height = 6}

g1 = ggplot(all_unique, aes(x = count, fill=type)) + 
   
  geom_histogram(binwidth = 1, alpha=0.7, position="identity") + 
   
  xlab("Unique values per feature (0-100)")+ theme(legend.position = "none") + 
   
  xlim(c(0,100)) +theme(axis.title.x=element_text(size=14,face="bold"))

g2 = ggplot(all_unique, aes(x = count, fill=type)) +  
   
  geom_histogram(binwidth = 100, alpha=0.7, position="identity") + 
   
  xlab("Unique values per feature(101+)")  + xlim(c(101,nrow(train))) +
   
  theme(axis.title.x=element_text(size=14, face="bold"))

grid.arrange(g1, g2, ncol=2)
```


LOOKING AT THE NUMBER OF NA`s per feature type

```{r  warning=FALSE, message=FALSE, fig.width = 8, fig.height = 6}

num_na = sapply(training_num, function(x) sum(is.na(x)))

str(lapply(training_num, function(x) sum(is.na(x))))

char_na = sapply(training_char, function(x) sum(is.na(x)))

str(lapply(training_char, function(x) sum(is.na(x))))

date_na = sapply(training_date, function(x) sum(is.na(x)))

str(lapply(training_date, function(x) sum(is.na(x))))

all_na = rbind(data.frame(count=num_na, type="Numerical"),
               
               data.frame(count=char_na, type="Character"), 
               
               data.frame(count=date_na, type="Date"))

all_na = data.frame(all_na)

all_na = all_na[all_na$count>0,]

head(all_na, 10)
```

PLOT OF NUMBER OF NA`s PER FEATURE TYPE

```{r  warning=FALSE, message=FALSE, fig.width = 8, fig.height = 6}

breaks <- c(5,10,50,100,500,1000,2000)

ggplot(all_na, aes(x = count, fill=type)) +  
  
  geom_histogram(alpha=0.7) + 

  # scale_y_log10(limits=c(1,2000), breaks=breaks) + 
  
  scale_x_log10(limits=c(1,20000), breaks=c(breaks,5000,10000,20000)) + 
  
  labs(title="Histogram of feature count per NA count", size=24, face="bold") +
  
  theme(plot.title = element_text(size = 16, face = "bold")) +
  
  xlab("NA Count") + ylab("Feature Count")

```

